Canonical rings of $\bq$-divisors on minimal rational surfaces

### 2016-07-24 surfaces-write-up/surface-write-up.tex
for merge
fixed a_i,b_i missing
minor edits to intro

Another closely related topic is the Hassett-Keel program \cite{hassett:classical-and-minimal-models}, which aims to describe log canonical models of the form 
\begin{align*}
	\overline {\mathscr M}_g(\alpha) := \bigoplus_{d \geq 0} u^d H^0 \left(
	\overline {\mathscr M}_g, \lfloor d K_{\overline{\mathscr M}_g} +
	\alpha\delta \rfloor  \right) 
\end{align*}
in terms of certain moduli spaces, where $\overline {\mathscr M_g}$ is the
moduli space of stable genus $g$ curves.
Moreover, $\bq$-divisors on surfaces also naturally
appear when considering the canonical ring of surfaces of Kodaira dimension 1.
A surface has Kodaira dimension 1 if and only if it is an elliptic surface \cite[p. 244]{
barthHPV:compactComplexSurfaces}, and
the canonical rings of elliptic surfaces most naturally described in the setting of
$\bq$-divisors \cite[Chapter V, Theorem 12.1]{barthHPV:compactComplexSurfaces}.

We expect the work in this paper will help calculate a presentation of certain rings of
Hilbert 
modular forms and Siegel modular forms. Recall that Hilbert and Siegel 
modular forms are two generalizations of modular forms to higher dimensions,
as described in \cite{geer:siegel-modular} and \cite{bruinier:hilbert-modular}.
Voight and Zureick-Brown show that 
one can translate the analytic category of 1-dimensional complex orbifolds to
the algebraic category of stacky curves \cite[Proposition 6.1.5]{vzb:stacky}. 
We expect that one can generalize this result to yield an equivalence of
categories between an appropriate subcategory of $m$-dimensional orbifolds and
a subcategory of $m$-dimensional algebraic spaces.
In particular, this would allow us to translate rings of Hilbert modular forms
and Siegel modular forms whose coarse space is a projective space or a
Hirzebruch surface to canonical rings on the corresponding algebraic surfaces.
Thereby, if our results could be extended to all rational surfaces,
they would give a bound on generators and relations for such
rings of Hilbert and Siegel modular forms parameterized by rational
Hilbert and Siegel modular surfaces.
This would be interesting because such modular surfaces tend to be
immensely complicated: ``Already for $g = 2$, the boundary of this fundamental domain is complicated; Gottschling found that it possesses $28$ boundary pieces'' \cite[p.\ 7]{geer:siegel-modular}.
Since these varieties are so complicated, it is difficult
to give explicit examples in which we can completely work out the canonical
ring of a modular surface.
Although we did not obtain a bound on the degree of generators and relations
for general rational surfaces, its enough work to get a result for the class of rational varieties we consider.

We will use $m \in \bz$ to index the dimension of a given projective
space $\bp^m$ or the type of the Hirzebruch surface $\hirz_m$.
Let $D$ be a (Weil) $\bq$-divisor of the form
\begin{align*}
	D = \sum_{i=1}^{n}\alpha_i D_i \in \bq \otimes_\bz \di X.
\end{align*}\noindent
where $n \in \bz$ indexes the number of irreducible divisors in 
the above expansion of $D$, $\alpha_i \in \bq$, and $D_i \in \di X$ is an integral
codimension 1 closed subscheme of $X$. When it is convenient to do
so, we shall sometimes start the indexing at $0$, so that $i$ runs
from $0$ to $n$. In the case $X = \bp^m$, define the degree of $D$ by $\deg D :=
\sum_{i=1}^{n}\alpha_i \cdot \deg D_i$. The floor of a $\bq$-divisor $D$ is a divisor $\lfloor D
\rfloor \in \di X$ is defined to be $\lfloor D \rfloor := \sum_{i = 1}^{n}
\lfloor \alpha_i \rfloor D_i$. Let $
	R_D := \bigoplus_{d \geq 0} u^d H^0(\bp^m, \lfloor dD \rfloor)$
denote the canonical ring associated to the $\bq$-divisor $D$. If $R$
is a graded ring, we denote the $d$th graded component of $R$ by $R_d$. If $r$ is a rational number, we let
$\fr(r) := r - \lfloor r
\rfloor$ denote the fractional part of $r$. 
Throughout, we work over a fixed algebraically closed field $\bk$.
Since generator and relation degrees are preserved under base change to the algebraic closure, assuming that $\bk$ is algebraically closed is no restriction
on generality and the results hold equally well over an arbitrary field.
If $D \in \bq \otimes_\bz
\di X$ is an arbitrary divisor, we denote $h^0(X, D) := \dim_\bk H^0(X,D)$.

@@

Then the canonical ring
\[
	R_D := \bigoplus_{d \geq 0} u^d H^0(\bp^m, \lfloor dD \rfloor)
\]

\noindent
is generated in degrees at most $\max_{0 \leq i \leq n}{k_i}$ with
relations generated in degrees at most $2 \max_{0 \leq i \leq n}{k_i}$.
\end{restatable}

@@

\begin{restatable}{thm}{restateproj}
\label{thm:proj-generators-relations}
Let $D = \sum_{i = 0}^n \alpha_i D_i \in \bq \otimes_\bz \di \bp^m$
with $\alpha_i = \frac{c_i}{k_i}\in \bq$ in reduced form. Write $\ell_i := \lcm
_{j \ne i}(k_j)$ and $a_i := \deg D_i$. Let $\bp^m \cong \proj
\bk[x_0, \ldots, x_m]$ and let each $D_i = V(f_i)$ with $f_i \in \bk[x_0, \ldots, 
x_m]$. Now suppose that $\{f_0, \ldots, f_n\}$ contains $m + 1$ independent 
linear polynomials in $x_0, \ldots, x_m$ {\rm(}with corresponding 
$\alpha_j$ possibly zero{\rm)}.

Then $R_D$ is generated in degrees at most 
\[
	\sum_{i = 0}^n \ell_i a_i
\]

\noindent
with relations generated in degrees at most
\[
	\max \left(2 \sum_{i=0}^n \ell_i a_i, \frac{\max_{0\le i \le n}
	(	\bida_i)}{\deg(D)} + \sum_{i=0}^n \ell_i a_i \right).
\]
\end{restatable}

@@

In Section \ref{sec:hirz}, we shift our attention to Hirzebruch surfaces. Recall that for each $m \geq 0$ we can define the
Hirzebruch surface $F_m := \proj(\sym^\bullet \sco_{\bp^1} \oplus
\sco_{\bp^1}(m)),$ or as a quotient of an open subscheme of
$\ba^4$ by a 
weighted action, as is done in the
beginning of Section \ref{sec:hirz}.
Note that Hirzebruch surfaces are regular, normal,
and locally factorial, being $\bp^1$-bundles over
$\bp^1$. Therefore, the natural map from Cartier
divisors to Weil divisors is an isomorphism.
In other words, all Cartier divisors arise as the zero locus
of a section of some line bundle. We implicitly use this in the
statement of the following theorem. The analogous fact for
projective space was also implicitly used in Theorems
~\ref{thm:proj-effective-intro} and ~\ref{thm:proj-generators-relations}.

\begin{thm}
\label{thm:hirz-generators-relations}
Let $D = \sum_{i = 1}^n \alpha_i D_i \in \bq
\otimes_\bz \di \hirz_m$ where $\alpha_i = \frac{c_i}{k_i} \in \bq$ is written in reduced form.
Let $u, v, z, w$ be the coordinates for the Hirzebruch surface
$\hirz_m$, as described at the beginning of Section
~\ref{sec:hirz} and suppose that $\{f_1, \ldots, f_n\}$
contains two independent 
linear polynomials in $u, v$ and two independent linear polynomials in
$w, x$ {\rm(}with corresponding $\alpha_j$ possibly zero{\rm)}.

Then $R_D$ is generated in degrees at most
\[
	\rho = \lcm_{1\le i\le n}(k_i)\sum_{1\le i\le j\le n} \bida_i \bidb_i
\]

\noindent
with relations generated in degrees at most 
\[
	2\rho
\]
\end{thm}

\begin{rem}
Theorem~\ref{thm:hirz-generators-relations} is restated with a more precise
bound in Theorem~\ref{thm:hirz-generators-relations-full}.
\end{rem}

@@

\begin{proof}
First, note that surjectivity of $\chi$ implies we have an exact sequence
\[
\begin{tikzcd}
0 \ar {r} & \ker \chi \ar{r}{\rho} & \ker \phi \ar {r}{\tau} & \ker \psi \ar r & 0.
\end{tikzcd}
\]

This shows that lifts of generators of $\ker \psi$ together with images of
generators of $\ker \chi$ generate all of $\ker \phi,$ as desired. 
\end{proof}

@@

\begin{thm}
\label{thm:hirz-generators-relations-full}
Let $D = \sum_{i = 1}^n \alpha_i D_i \in \bq
\otimes_\bz \di \hirz_m$ where $\alpha_i = \frac{c_i}{k_i} \in \bq$ is written in reduced form.
Write $\ell_{i,j} := \lcm_{h\ne i,j}(k_h)$.
Let $u, v, z, w$ be the coordinates for the Hirzebruch surface
$\hirz_m$, as described at the beginning of Section
~\ref{sec:hirz} and suppose that $\{f_1, \ldots, f_n\}$
contains two independent 
linear polynomials in $u, v$ and two independent linear polynomials in
$w, x$ {\rm(}with corresponding $\alpha_j$ possibly zero{\rm)}.
Define $\Te, \Tp$, and $\Tm$ as in Definition ~\ref{defn:t-defn}. Let each $D_i = V(f_i)$ where $f_i \in \sco(a_i, b_i)$.

@@

\noindent
with relations generated in degrees at most 
\[
	2\rho
\]
\end{thm}
\begin{rem}
This implies theorem \ref{thm:hirz-generators-relations}.
\end{rem}
\begin{proof}
The generation degree bound is as stated in Lemma \ref{lem:hirz-generators}.
By Proposition \ref{lem:composite-map}, the degree of generation of
$\ker \phi$ is at most the maximum of the generation degrees of $\ker \chi$
and $\ker \psi$, giving us the desired relations bound. The bound on $\ker \chi$
follows from Lemma ~\ref{lem:bound-ker-chi} and the bound on $\ker \psi$
follows from Lemma ~\ref{lem:hirz-bound-ker-psi}.
\end{proof}

>>>>

Another closely related topic is the Hassett-Keel program \cite{hassett:classical-and-minimal-models}, which aims to describe log canonical models of the form 
\begin{align*}
	\overline {\mathscr M}_g(\alpha) := \bigoplus_{d \geq 0} u^d H^0 \left(
	\overline {\mathscr M}_g, \lfloor d K_{\overline{\mathscr M}_g} +
	\alpha\delta \rfloor  \right) 
\end{align*}
in terms of certain moduli spaces, where $\overline {\mathscr M_g}$ is the
moduli space of stable genus $g$ curves.
Moreover, $\bq$-divisors on surfaces also naturally
appear when considering the canonical ring of surfaces of Kodaira dimension 1.
A surface has Kodaira dimension 1 if and only if it is an elliptic surface \cite[p. 244]{
barthHPV:compactComplexSurfaces}, and
the canonical rings of elliptic surfaces are most naturally described in the setting of
$\bq$-divisors \cite[Chapter V, Theorem 12.1]{barthHPV:compactComplexSurfaces}.

We expect the work in this paper will help calculate a presentation of certain rings of
Hilbert 
modular forms and Siegel modular forms. Recall that Hilbert and Siegel 
modular forms are two generalizations of modular forms to higher dimensions,
as described in \cite{geer:siegel-modular} and \cite{bruinier:hilbert-modular}.
Voight and Zureick-Brown show that 
one can translate the analytic category of 1-dimensional complex orbifolds to
the algebraic category of stacky curves \cite[Proposition 6.1.5]{vzb:stacky}. 
We expect that one can generalize this result to yield an equivalence of
categories between an appropriate subcategory of $m$-dimensional orbifolds and
a subcategory of $m$-dimensional algebraic spaces.
In particular, this would allow us to translate rings of Hilbert modular forms
and Siegel modular forms whose coarse space is a projective space or a
Hirzebruch surface to canonical rings on the corresponding algebraic surfaces.
Thereby, if our results could be extended to all rational surfaces,
they would give a bound on generators and relations for such
rings of Hilbert and Siegel modular forms parameterized by rational
Hilbert and Siegel modular surfaces.
This would be interesting because such modular surfaces tend to be
immensely complicated: ``Already for $g = 2$, the boundary of this fundamental domain is complicated; Gottschling found that it possesses $28$ boundary pieces'' \cite[p.\ 7]{geer:siegel-modular}.
Since these varieties are so complicated, it is difficult
to give explicit examples in which we can completely work out the canonical
ring of a modular surface.
Although we did not obtain a bound on the degree of generators and relations
for general rational surfaces, the restricted class of rational varieties we consider still required significant work.

\subsection{Notation}
We now define some notation used throughout the paper.
Use $m \in \bz$ to index the dimension of a given projective
space $\bp^m$ and the type of the Hirzebruch surface $\hirz_m$.
Let $D$ be a (Weil) $\bq$-divisor of the form
\begin{align*}
	D = \sum_{i=1}^{n}\alpha_i D_i \in \bq \otimes_\bz \di X.
\end{align*}\noindent
where $n \in \bz$ indexes the number of irreducible divisors in 
the above expansion of $D$, $\alpha_i \in \bq$, and $D_i \in \di X$ is an integral
codimension 1 closed subscheme of $X$. When it is convenient to do
so, we shall sometimes start the indexing at $0$, so that $i$ runs
from $0$ to $n$. In the case $X = \bp^m$, define the degree of $D$ by $\deg D :=
\sum_{i=1}^{n}\alpha_i \cdot \deg D_i$. The floor of a $\bq$-divisor $D$ is a divisor $\lfloor D
\rfloor \in \di X$ is defined to be $\lfloor D \rfloor := \sum_{i = 1}^{n}
\lfloor \alpha_i \rfloor D_i$. Let 
$R_D := \bigoplus_{d \geq 0} u^d H^0(\bp^m, \lfloor dD \rfloor)$
denote the canonical ring associated to the $\bq$-divisor $D$. 
So, $R_D$ is notation for $R(X,D)$ when $X$ is understood from context.
If $R$
is a graded ring, we denote the $d$th graded component of $R$ by $R_d$. If $r$ is a rational number, we let
$\fr(r) := r - \lfloor r
\rfloor$ denote the fractional part of $r$. 
Throughout, we work over a fixed algebraically closed field $\bk$.
Since generator and relation degrees are preserved under base change to the algebraic closure, assuming that $\bk$ is algebraically closed is no restriction
on generality and the results hold equally well over an arbitrary field.
If $D \in \bq \otimes_\bz
\di X$ is an arbitrary divisor, we denote $h^0(X, D) := \dim_\bk H^0(X,D)$.

@@

Then the canonical ring $R_D$
is generated in degrees at most $\max_{0 \leq i \leq n}{k_i}$ with
relations generated in degrees at most $2 \max_{0 \leq i \leq n}{k_i}$.
\end{restatable}

@@

\begin{restatable}{thm}{restateproj}
\label{thm:proj-generators-relations}
Let $D = \sum_{i = 0}^n \alpha_i D_i \in \bq \otimes_\bz \di \bp^m$
with $\alpha_i = \frac{c_i}{k_i}\in \bq$ in reduced form. Write $\ell_i := \lcm
_{j \ne i}(k_j)$ and $a_i := \deg D_i$. Let $\bp^m \cong \proj
\bk[x_0, \ldots, x_m]$ and define $D_i := V(f_i)$, for some $f_i \in \bk[x_0, \ldots, 
x_m]$. Now suppose that $\{f_0, \ldots, f_n\}$ contains $m + 1$ independent 
linear polynomials in $x_0, \ldots, x_m$ {\rm(}where we allow for 
$\alpha_j = 0$ in the expansion of $D${\rm)}.

Then $R_D$ is generated in degrees at most 
$\sum_{i = 0}^n \ell_i a_i$
with relations generated in degrees at most
\[
	\max \left(2 \sum_{i=0}^n \ell_i a_i, \frac{\max_{0\le i \le n}
	(	\bida_i)}{\deg(D)} + \sum_{i=0}^n \ell_i a_i \right).
\]
\end{restatable}

@@

In Section \ref{sec:hirz}, we shift our attention to Hirzebruch surfaces. Recall that for each $m \geq 0$ we can define the
Hirzebruch surface $F_m := \proj(\sym^\bullet \sco_{\bp^1} \oplus
\sco_{\bp^1}(m))$ as a projective bundle over $\bp^1$.
Let $u,v$ the projective coordinates on the base $\bp^1$ and let $z,w$ be the projective coordinates
on the fiber, as defined more precisely at the beginning of
\ref{sec:hirz}.
Since Hirzebruch surfaces are smooth, the natural map from Cartier
divisors to Weil divisors is an isomorphism.
In other words, all Cartier divisors arise as the zero locus
of a section of some line bundle. We implicitly use this in the
statement of the following theorem. The analogous fact for
projective space was also implicitly used in Theorems
~\ref{thm:proj-effective-intro} and ~\ref{thm:proj-generators-relations}.

\begin{thm}
\label{thm:hirz-generators-relations}
Let $D = \sum_{i = 1}^n \alpha_i D_i \in \bq
\otimes_\bz \di \hirz_m$ where $\alpha_i = \frac{c_i}{k_i} \in \bq$ is written in reduced form.  
Let each $D_i = V(f_i)$, where $f_i \in \sco(a_i, b_i)$.
Let $u, v, z, w$ be the coordinates for the Hirzebruch surface
$\hirz_m$, and suppose that $\{f_1, \ldots, f_n\}$
contains two independent 
linear polynomials in $u, v$ and two independent linear polynomials in
$w, x$ {\rm(}where we allow for $\alpha_j = 0$ in the expansion of $D${\rm)}.

Then $R_D$ is generated in degrees at most
\[
	\lcm_{1\le i\le n}(k_i)\cdot \left(\sum_{1\le i\le j\le n} \bida_i \bidb_i\right) 
\]
\noindent
with relations generated in degrees at most 
\[
	2 \cdot \lcm_{1\le i\le n}(k_i) \cdot \left( \sum_{1\le i\le j\le n} \bida_i \bidb_i\right) 
\]
\end{thm}
\begin{rem}
Theorem~\ref{thm:hirz-generators-relations} is restated with a more precise
bound in Theorem~\ref{thm:hirz-generators-relations-full}.
\end{rem}

@@

\begin{proof}
First, note that surjectivity of $\chi$ implies we have an exact sequence
\[
\begin{tikzcd}
0 \ar {r} & \ker \chi \ar{r} & \ker \phi \ar {r} & \ker \psi \ar r & 0.
\end{tikzcd}
\]
This shows that lifts of generators of $\ker \psi$ together with images of
generators of $\ker \chi$ generate all of $\ker \phi,$ as desired. 
\end{proof}

@@

\begin{thm}
\label{thm:hirz-generators-relations-full}
Let $D = \sum_{i = 1}^n \alpha_i D_i \in \bq
\otimes_\bz \di \hirz_m$ where $\alpha_i = \frac{c_i}{k_i} \in \bq$ is written in reduced form.
Write $\ell_{i,j} := \lcm_{h\ne i,j}(k_h)$.
Let $u, v, z, w$ be the coordinates for the Hirzebruch surface
$\hirz_m$, as described at the beginning of Section
~\ref{sec:hirz} and suppose that $\{f_1, \ldots, f_n\}$
contains two independent 
linear polynomials in $u, v$ and two independent linear polynomials in
$w, x$ {\rm(}with corresponding $\alpha_j$ possibly zero{\rm)}.
Recall $\Te, \Tp$, and $\Tm$ as given in Definition ~\ref{defn:t-defn} and let each $D_i = V(f_i)$ where $f_i \in \sco(a_i, b_i)$.

@@

\noindent
with relations generated in degrees at most $2 \rho$.
\end{thm}
%\begin{rem}
%This implies theorem \ref{thm:hirz-generators-relations}.
%\end{rem}
\begin{proof}
The generation degree bound is as stated in Lemma \ref{lem:hirz-generators}.
By Proposition \ref{lem:composite-map}, the degree of generation of
$\ker \phi$ is at most the maximum of the generation degrees of $\ker \chi$
and $\ker \psi$, giving us the desired relations bound. The bound on $\ker \chi$
follows from Lemma ~\ref{lem:bound-ker-chi} and the bound on $\ker \psi$
follows from Lemma ~\ref{lem:hirz-bound-ker-psi}.
\end{proof}
